---
title: "Calibration Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibration Guide}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(hydroMetrics)
```

This vignette shows how to choose and use `hydroMetrics` metrics in practical
calibration workflows. The focus is narrow: deterministic calibration guidance
for the current package surface, not a general review of hydrologic calibration
theory.

In `hydroMetrics`, calibration means comparing simulated and observed series
with metrics that emphasize different aspects of fit such as overall error,
correlation, bias, variability, seasonality, or selected flow regimes. No
single metric is universally best. A calibration target should match the
modeling objective.

This guide uses the current public contract:

- canonical metric selection through `gof()` and discovery-oriented helpers
- compatibility wrappers where a retained historical entry point is useful
- deprecated forwarding wrappers only as transitional aliases, not as primary
  current metric names

## Start With the Recommended Shortlist

For most deterministic calibration workflows, the safest starting point is the
same compat-10 shortlist that powers the default `gof()` behavior:

- `nse`
- `kge`
- `rmse`
- `pbias`
- `mae`
- `mse`
- `r2`
- `ve`
- `rsr`
- `nrmse`

The package currently exposes that curated shortlist through the internal
public discovery helpers rather than through internal registry access. The main
public user path is simply to call `gof(sim, obs)` with no explicit `methods`,
but `metric_search()` also makes the current recommended shortlist visible.

```{r}
metric_search(preset = "recommended")[, c("id", "name")]
```

## Core Deterministic Calibration Metrics

The public default is `gof(sim, obs)`. It returns the compat-10 metrics on
aligned prepared data.

```{r}
sim <- c(1.2, 1.8, 3.4, 3.9, 5.1, 6.0, 7.2, 8.1)
obs <- c(1.0, 2.0, 3.0, 4.0, 5.0, 6.2, 7.0, 8.0)

default_out <- gof(sim, obs)
default_out
```

These metrics emphasize different calibration properties:

- `nse`, `kge`: overall efficiency relative to observed behavior
- `rmse`, `mae`, `mse`, `nrmse`: absolute or squared error magnitude
- `pbias`: directional volume bias
- `r2`: linear association strength
- `ve`: volumetric agreement
- `rsr`: error relative to observed variability

Two package-specific wording details matter here:

- `pbias` uses `100 * sum(sim - obs) / sum(obs)`, so positive values indicate
  overestimation. Moriasi et al. (2007) uses the opposite-sign interpretation,
  so its qualitative PBIAS thresholds should not be transferred verbatim without
  sign-aware translation.
- `nrmse` is the package's mean(obs)-normalized variant; published NRMSE
  definitions vary, so the exact normalization should be named explicitly in
  downstream reporting.

When you want a single metric directly, the exported wrappers are often more
convenient than a multi-metric `gof()` call. For core metrics that are not
exposed as direct wrappers in the current public namespace, use
`gof(methods = ...)`. The current compatibility-oriented exported wrappers are
`HFB()`, `NSeff()`, `alpha()`, `beta()`, `mae()`, `mNSeff()`,
`pbias()`, `r()`, `rNSeff()`, `rsr()`, and `wsNSeff()`. Core metrics such as
`kge`, `rmse`, `r2`, and `nrmse` are available through `gof()` / `ggof()`
method selection rather than standalone exported wrapper functions.

```{r}
gof(sim, obs, methods = c("nse", "kge", "rmse"))
mae(sim, obs)
pbias(sim, obs)
```

## When to Use `gof()` vs Wrappers

Use `gof()` when you want one consistent preprocessing path and several metrics
at once:

```{r}
gof(sim, obs, methods = c("nse", "kge", "rmse", "pbias"))
```

Use a wrapper when the metric itself is the focus or when the metric has a
more specialized contract that you want to call explicitly. For current
canonical naming, prefer the canonical exported wrapper when one exists:

```{r}
composite_performance_index(sim, obs)
```

The deprecated forwarding wrapper `extended_valindex()` still resolves to
`composite_performance_index()`, but it is transitional rather than the
preferred current name.

`valindex()` is also available, but it remains the historical orchestration
wrapper around `gof(methods = fun, ...)`, not a fixed composite score:

```{r}
valindex(sim, obs, fun = c("nse", "rmse"))
```

## Extended Deterministic Metrics

`gof(sim, obs, extended = TRUE)` expands omitted method selection from the
compat-10 default to all registered metrics that are applicable to the current
input context.

This does not mean every metric is always included. Conditional metrics are
only auto-selected when their documented preconditions are satisfied.

```{r}
ext_out <- gof(sim, obs, extended = TRUE)
sort(intersect(
  c(
    "entropy_diff",
    "upper_tail_conditional_exceedance",
    "composite_performance_index"
  ),
  names(ext_out)
))
```

For a plain numeric vector, `extended = TRUE` can expose useful additional
diagnostics such as distribution, information-theory, or composite metrics, but
it still respects conditional gating.

## Seasonal, Regime, and Tail-Sensitive Metrics

Some calibration targets are about more than overall fit. If the modeling goal
is seasonal timing, regime shape, or selected extremes, metric choice should
reflect that.

### Low-Flow Diagnostics

Metrics such as `low_flow_bias` are useful when calibration is aimed at low-flow
behavior instead of only pointwise error. In `hydroMetrics`, `low_flow_bias` is
the package's observed lower-30% subset percent-bias diagnostic rather than a
claim of literature-exact low-flow bias equivalence.

These can be requested directly when that is the modeling objective:

```{r}
gof(sim, obs, methods = c("low_flow_bias"))
```

### Tail and Event-Sensitive Metrics

Metrics such as `upper_tail_conditional_exceedance`, `extreme_event_ratio`, and
`event_nse` are narrower. They rely on observed tail or event structure. When
that structure is absent, `gof(extended = TRUE)` excludes them rather than
failing.

```{r}
sim_flat <- c(1, 2, 2, 2, 2, 2)
obs_flat <- c(1, 1, 1, 1, 2, 2)

intersect(
  c("upper_tail_conditional_exceedance", "extreme_event_ratio", "event_nse"),
  names(gof(sim_flat, obs_flat, extended = TRUE))
)
```

The deprecated forwarding wrapper `tail_dependence_score()` remains available
for transition, but the canonical current metric name is
`upper_tail_conditional_exceedance`.

These are best treated as targeted diagnostics, not generic first-choice
calibration objectives.

## Practical Calibration Workflows

### A conservative deterministic workflow

Start with the default shortlist, review a few complementary metrics, then add
specialized diagnostics only if the modeling objective requires them.

```{r}
default_out[c("nse", "kge", "rmse", "pbias", "ve", "rsr")]
```

If you want one fixed scalar summary for comparing candidate runs, the package
also provides `composite_performance_index()`. It is a package-defined
composite of stable core metrics, so it can be useful as a secondary summary,
but it should not replace objective-specific diagnostics.

```{r}
composite_performance_index(sim, obs)
```

The deprecated forwarding wrapper `extended_valindex()` still routes to the same
canonical metric, but `composite_performance_index()` is the preferred current
name.

## Caveats for Metric Choice

- No single metric is universally best for calibration.
- Default `gof()` is a stable starting point because it uses the curated
  compat-10 shortlist.
- `gof(extended = TRUE)` is useful for exploration, but conditional metrics
  appear only when their contracts are satisfied.
- Tail/event metrics require qualifying observed extremes or events.
- Log-sensitive or zero-sensitive metrics should only be used when their input
  assumptions match the data and calibration objective.

Choose metrics that match the behavior you want the model to reproduce:
overall efficiency, low-flow fidelity, bias, regime shape, or extreme-event
behavior.

## Next Steps

Use this vignette as the calibration-oriented starting point, then consult:

- `vignette("getting-started", package = "hydroMetrics")` for a minimal package
  walkthrough
- `?gof`, `?ggof`, and wrapper help pages for exact function contracts
- [NEWS.md](../NEWS.md) for the current Phase 3 implementation boundary and
  closure notes
