<!--
%\VignetteIndexEntry{Getting Started with hydroMetrics}
%\VignetteEngine{knitr::knitr}
%\VignetteEncoding{UTF-8}
-->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`hydroMetrics` provides clean-room hydrological model-evaluation metrics with a
small orchestration layer for deterministic preprocessing and wrapper
compatibility.

## Installation

For local release validation, build the source tarball from the package root
with `R CMD build .` and install the generated bundle:

```r
install.packages("hydroMetrics_0.4.0.tar.gz", repos = NULL, type = "source")
```

## A small discovery and evaluation workflow

The current package surface is built around:

- `preproc()` for deterministic alignment and NA handling
- `metric_search()` for light-weight metric discovery
- `gof()` for orchestration over one or more selected metrics
- compatibility wrappers such as `NSeff()` when a single historical entry point
  is more convenient

```{r}
library(hydroMetrics)

sim_raw <- c(1.05, 1.80, NA, 3.60, 4.10, 5.00)
obs_raw <- c(1.00, 2.00, 2.80, 3.40, 4.00, 5.20)
```

## Optional preprocessing

`preproc()` makes the alignment and NA policy explicit before metric
evaluation. Here we remove the incomplete pair and keep the processed vectors.

```{r}
prep <- preproc(sim_raw, obs_raw, na_strategy = "remove")
prep
```

The processed vectors are then available as `prep$sim` and `prep$obs`.

## Discover a modest metric subset

`metric_search()` returns a simple `data.frame` with registry metadata plus
discovery annotations such as exported wrapper names, compatibility-export
status, and preset membership.

```{r}
recommended <- metric_search(preset = "recommended", exported = TRUE)
recommended[, c("id", "name", "exported_wrappers", "presets")]
```

For a short deterministic summary, select a few metrics from the recommended
set.

```{r}
selected_ids <- recommended$id[recommended$id %in% c("nse", "rmse", "pbias")]
selected_ids
```

## Evaluate the selected metrics

The primary orchestration entry point is `gof()`. For single-series inputs it
returns a named numeric vector with class `hydro_metrics`.

```{r}
gof_out <- gof(prep$sim, prep$obs, methods = selected_ids)
gof_out
class(gof_out)
attr(gof_out, "n_obs")
```

The metric names are the vector names, and the `n_obs` attribute records the
number of paired values used after preprocessing.

## Compatibility-oriented path

If you only need one metric, a compatibility wrapper can still be more
convenient.

```{r}
NSeff(prep$sim, prep$obs)
```

Uppercase hydroGOF-style names are also accepted as orchestration labels inside
`gof()`, even though they are not exported standalone functions.

```{r}
gof(prep$sim, prep$obs, methods = "NSE")
```

## Reading the workflow

- Use `metric_search()` when you want to discover relevant metrics by preset,
  tags, text, or compatibility status.
- Use `gof()` when you want one orchestrated call over a selected metric set.
- Use an exported compatibility wrapper such as `NSeff()` when you want a
  single historical entry point with current package behavior.

- Uppercase hydroGOF names such as `NSE`, `KGE`, `RMSE`, `R2`, `NRMSE`, and
  `PBIAS` are accepted in `gof()` / `ggof()` `methods`, but they are not
  exported standalone functions in the package namespace.
- `ggof()` is tabular output only; it does not create plots.
- `preproc()` currently supports single-series vector or indexed inputs only.
- Formal compatibility aliases such as `na.rm`, `fun`, `keep`,
  `epsilon.type`, and `epsilon.value` are documented where already supported by
  the package.
