<!--
%\VignetteIndexEntry{Getting Started with hydroMetrics}
%\VignetteEngine{knitr::knitr}
%\VignetteEncoding{UTF-8}
-->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`hydroMetrics` provides clean-room hydrological model-evaluation metrics with a
small orchestration layer for deterministic preprocessing and wrapper
compatibility.

## Installation

For local release validation, build the source tarball from the package root
with `R CMD build .` and install the generated bundle:

```r
install.packages("hydroMetrics_0.2.0.tar.gz", repos = NULL, type = "source")
```

## Core usage

The primary entry point is `gof()`. It evaluates one or more registered metrics
over aligned simulated and observed series.

```{r}
library(hydroMetrics)

sim <- c(1, 2, 3, 4)
obs <- c(1, 2, 2, 4)

gof_out <- gof(sim, obs, methods = c("NSE", "rmse", "pbias"))
gof_out
```

## Batch summaries with `ggof()`

`ggof()` returns a tabular `hydro_metrics_batch` object and does not open a
graphics device.

```{r}
ggof(sim, obs, methods = c("NSE", "rmse"))
```

## Wrapper examples

Compatibility wrappers remain available when a single metric is more
convenient.

```{r}
NSeff(sim, obs)
mae(sim, obs)
pbias(sim, obs)
```

## Compatibility notes

- `ggof()` is tabular output only; it does not create plots.
- `APFB()` requires indexed `zoo` or `xts` inputs.
- `preproc()` currently supports single-series vector or indexed inputs only.
- Formal compatibility aliases such as `na.rm`, `fun`, `keep`,
  `epsilon.type`, and `epsilon.value` are documented where already supported by
  the package.
