---
title: "Metric Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metric Reference}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(hydroMetrics)
```

This vignette is a concise catalog of the current `hydroMetrics` metric
surface. It is intended as a navigation guide: where metrics live, how they
are grouped, and which entry point to use for each class of metric.

For worked examples and metric-selection advice, use the dedicated workflow
guides instead of treating this document as a full tutorial.

## Overview

The package exposes metrics through three main routes:

- `gof(sim, obs)` for the stable default deterministic shortlist
- `gof(sim, obs, extended = TRUE)` for the broader deterministic registry
  surface when the current input satisfies each metric's contract
- direct wrappers for metrics that are either convenient one-off calls or
  require a more specialized input shape

The default `gof()` behavior is intentionally conservative. It stays aligned to
the compat-10 deterministic shortlist, while `extended = TRUE` expands to the
broader registered deterministic metric set that can safely run on the current
input.

Discovery-facing metric IDs are the lowercase canonical registry IDs returned
by helpers such as `metric_search()` and `metric_preset()`. Compatibility
wrappers and deprecated forwarding wrappers may still be exported, but they are
not independent canonical discovery IDs.

```{r}
sim <- c(1.2, 1.8, 3.4, 3.9, 5.1, 6.0, 7.2, 8.1)
obs <- c(1.0, 2.0, 3.0, 4.0, 5.0, 6.2, 7.0, 8.0)

names(gof(sim, obs))
```

## Discovering Metrics

For most users, the easiest discovery path is to start with `gof()` for the
default deterministic shortlist, then use the public discovery helpers when a
more targeted metric set is needed.

`metric_preset("recommended")` returns the curated recommended shortlist that
matches the stable default deterministic starting point.

```{r}
metric_preset("recommended")
```

`metric_search()` provides a lightweight public discovery view over the current
registry metadata. It can be used to search by preset, category, tags, and
text without relying on internal helpers.

```{r}
metric_search(preset = "recommended")[, c("id", "name", "presets")]
```

For broader discovery, search by category or text:

```{r}
head(metric_search(category = "correlation")[, c("id", "name", "category")], 12)
```

## Exported Wrapper Naming

The current exported wrapper surface contains three distinct kinds of names:

- canonical exported wrappers such as `alpha()`, `beta()`, `mae()`, `pbias()`,
  `r()`, and `rsr()`
- retained compatibility exports such as `HFB()`, `NSeff()`, `mNSeff()`,
  `rNSeff()`, `wsNSeff()`, and `mutual_information_score()`
- deprecated forwarding wrappers such as `tail_dependence_score()` and
  `extended_valindex()`, which remain callable temporarily but are no longer
  canonical current metric names

Uppercase hydroGOF names such as `NSE`, `KGE`, `RMSE`, `R2`, `NRMSE`, and
`PBIAS` remain available as `gof()` / `ggof()` method labels, not as exported
standalone functions.

## Default Deterministic Metrics

With no explicit `methods`, `gof(sim, obs)` evaluates the recommended
deterministic shortlist. These are the package's most compatibility-stable
starting metrics for everyday evaluation and calibration.

They cover the main deterministic fit dimensions:

- efficiency and overall agreement: `nse`, `kge`, `ve`
- error magnitude: `rmse`, `mae`, `mse`, `nrmse`
- bias and association: `pbias`, `r2`
- variability-relative error: `rsr`

Use these when the goal is broad deterministic model evaluation without a more
specialized structural target.

## Extended Deterministic Metrics

`gof(sim, obs, extended = TRUE)` can expose many additional deterministic
metrics when the current input context supports them.

These include Layer A, Layer B, and Layer C additions such as:

- robust and alternative loss metrics: `huber_loss`, `quantile_loss`,
  `trimmed_rmse`, `winsor_rmse`
- hydrograph and event diagnostics: `peak_timing_error`,
  `rising_limb_error`, `event_nse`
- distribution and information-theory diagnostics: `entropy_diff`,
  `mutual_information`, `normalised_mi`
- composite and summary diagnostics: `composite_performance_index`,
  `rank_turnover_score`, `distribution_overlap`, `quantile_shift_index`

This route is broader, but not unconditional. Auto-selection remains gated by
each metric's documented contract.

```{r}
sort(intersect(
  c(
    "entropy_diff",
    "mutual_information",
    "composite_performance_index"
  ),
  names(gof(sim, obs, extended = TRUE))
))
```

The deprecated forwarding wrapper `extended_valindex()` still routes to
`composite_performance_index()`, but the canonical current metric name is
`composite_performance_index`.

## Tail, Event, and Regime-Sensitive Metrics

Some metrics are deterministic but still conditional because they require
observed structure such as qualifying tails, events, or hydrograph phases.

Examples include:

- `upper_tail_conditional_exceedance`
- `extreme_event_ratio`
- `event_nse`
- `rising_limb_error`
- `recession_constant`

These metrics are available through the registry and public wrappers where
exported, but `gof(extended = TRUE)` includes them only when the observed data
meet their documented preconditions.

```{r}
intersect(
  c("upper_tail_conditional_exceedance", "extreme_event_ratio", "event_nse"),
  names(gof(sim, obs, extended = TRUE))
)
```

The deprecated forwarding wrapper `tail_dependence_score()` still routes to
`upper_tail_conditional_exceedance()`, but it is transitional rather than the
preferred current name.

## Where To Go Next

Use this vignette as the package map, then move to the workflow guide that
matches your task:

- `vignette("getting-started", package = "hydroMetrics")` for a minimal
  package walkthrough
- `vignette("calibration-guide", package = "hydroMetrics")` for deterministic
  calibration workflows and metric selection

For exact function contracts, consult `?gof`, `?ggof`, and the relevant metric
wrapper help pages.
