---
title: "Metric Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metric Reference}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(hydroMetrics)
```

This vignette is a concise catalog of the current `hydroMetrics` metric
surface. It is intended as a navigation guide: where metrics live, how they
are grouped, and which entry point to use for each class of metric.

For worked examples and metric-selection advice, use the dedicated workflow
guides instead of treating this document as a full tutorial.

## Overview

The package exposes metrics through three main routes:

- `gof(sim, obs)` for the stable default deterministic shortlist
- `gof(sim, obs, extended = TRUE)` for the broader deterministic registry
  surface when the current input satisfies each metric's contract
- direct wrappers for metrics that are either convenient one-off calls or
  require a more specialized input shape

The default `gof()` behavior is intentionally conservative. It stays aligned to
the compat-10 deterministic shortlist, while `extended = TRUE` expands to the
broader registered deterministic metric set that can safely run on the current
input.

```{r}
sim <- c(1.2, 1.8, 3.4, 3.9, 5.1, 6.0, 7.2, 8.1)
obs <- c(1.0, 2.0, 3.0, 4.0, 5.0, 6.2, 7.0, 8.0)

names(gof(sim, obs))
```

## Discovering Metrics

For most users, the easiest discovery path is to start with `gof()` and then
move to explicit wrappers or explicit `methods = ...` requests as needed.

Internally, the package also maintains a registry listing. The internal helper
`hydroMetrics:::list_metrics(recommended = TRUE)` returns the curated
recommended shortlist that matches the stable default deterministic starting
point.

```{r}
hydroMetrics:::list_metrics(recommended = TRUE)[, c("id", "name")]
```

The full internal registry is also available for audit and discovery work:

```{r}
head(hydroMetrics:::list_metrics()[, c("id", "name", "tags")], 12)
```

## Exported Wrapper Naming

The current compatibility-oriented wrapper names exported from the package
namespace are:

- `HFB()`
- `NSeff()`
- `alpha()`
- `beta()`
- `mae()`
- `mNSeff()`
- `pbias()`
- `r()`
- `rNSeff()`
- `rsr()`
- `wsNSeff()`

Uppercase hydroGOF names such as `NSE`, `KGE`, `RMSE`, `R2`, `NRMSE`, and
`PBIAS` remain available as `gof()` / `ggof()` method labels, not as exported
standalone functions.

## Default Deterministic Metrics

With no explicit `methods`, `gof(sim, obs)` evaluates the recommended
deterministic shortlist. These are the package's most compatibility-stable
starting metrics for everyday evaluation and calibration.

They cover the main deterministic fit dimensions:

- efficiency and overall agreement: `nse`, `kge`, `ve`
- error magnitude: `rmse`, `mae`, `mse`, `nrmse`
- bias and association: `pbias`, `r2`
- variability-relative error: `rsr`

Use these when the goal is broad deterministic model evaluation without a more
specialized structural target.

## Extended Deterministic Metrics

`gof(sim, obs, extended = TRUE)` can expose many additional deterministic
metrics when the current input context supports them.

These include Layer A, Layer B, and Layer C additions such as:

- robust and alternative loss metrics: `huber_loss`, `quantile_loss`,
  `trimmed_rmse`, `winsor_rmse`
- flow-duration and regime diagnostics: `fdc_slope_error`,
  `fdc_highflow_bias`, `fdc_lowflow_bias`, `fdc_shape_distance`,
  `flow_duration_entropy`
- hydrograph and event diagnostics: `peak_timing_error`,
  `rising_limb_error`, `event_nse`
- distribution and information-theory diagnostics: `entropy_diff`,
  `mutual_information`, `normalised_mi`
- composite and summary diagnostics: `extended_valindex`,
  `rank_turnover_score`, `distribution_overlap`, `quantile_shift_index`

This route is broader, but not unconditional. Auto-selection remains gated by
each metric's documented contract.

```{r}
sort(intersect(
  c(
    "entropy_diff",
    "mutual_information",
    "extended_valindex",
    "flow_duration_entropy"
  ),
  names(gof(sim, obs, extended = TRUE))
))
```

## Tail, Event, and Regime-Sensitive Metrics

Some metrics are deterministic but still conditional because they require
observed structure such as qualifying tails, events, or hydrograph phases.

Examples include:

- `tail_dependence_score`
- `extreme_event_ratio`
- `event_nse`
- `rising_limb_error`
- `recession_constant`

These metrics are available through the registry and public wrappers where
exported, but `gof(extended = TRUE)` includes them only when the observed data
meet their documented preconditions.

```{r}
intersect(
  c("tail_dependence_score", "extreme_event_ratio", "event_nse"),
  names(gof(sim, obs, extended = TRUE))
)
```

## Where To Go Next

Use this vignette as the package map, then move to the workflow guide that
matches your task:

- `vignette("getting-started", package = "hydroMetrics")` for a minimal
  package walkthrough
- `vignette("calibration-guide", package = "hydroMetrics")` for deterministic
  calibration workflows and metric selection

For exact function contracts, consult `?gof`, `?ggof`, and the relevant metric
wrapper help pages.
