---
title: "Uncertainty and Probabilistic Evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uncertainty and Probabilistic Evaluation}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(hydroMetrics)
```

This vignette shows how to use the current `hydroMetrics` uncertainty and
probabilistic evaluation tools. The scope is narrow: interval, ensemble, and
score-against-baseline evaluation with the package surface that exists today.

In `hydroMetrics`, deterministic fit metrics and probabilistic metrics answer
different questions. A deterministic score asks how close one simulated series
is to one observed series. Probabilistic metrics instead evaluate forecast
distributions, predictive intervals, or relative improvement against a
reference score.

## What hydroMetrics Supports Today

The current uncertainty-facing metrics are explicit wrappers:

- `crps()` for ensemble forecasts
- `picp()` for prediction-interval coverage
- `mwpi()` for mean prediction-interval width
- `skill_score()` for lower-is-better relative improvement against a baseline

These are not part of the default deterministic `gof(sim, obs)` workflow, and
they are not auto-selected by `gof(sim, obs, extended = TRUE)` for ordinary
single simulated and observed vectors.

```{r}
sim <- c(1.2, 1.8, 3.4, 3.9, 5.1, 6.0)
obs <- c(1.0, 2.0, 3.0, 4.0, 5.0, 6.2)

names(gof(sim, obs))
intersect(c("crps", "picp", "mwpi", "skill_score"), names(gof(sim, obs, extended = TRUE)))
```

That boundary is intentional. These metrics require shaped inputs that do not
fit the plain deterministic `gof(sim, obs)` contract.

## Input Contracts and Data Shape

### Ensemble Evaluation with `crps()`

`crps()` expects:

- a numeric matrix `sim`
- rows = forecast cases
- columns = ensemble members
- `obs` as a numeric vector with length `nrow(sim)`

```{r}
ens <- matrix(
  c(
    1.0, 1.2, 0.8,
    2.0, 2.2, 1.8,
    3.0, 3.2, 2.8
  ),
  nrow = 3,
  byrow = TRUE
)
obs_ens <- c(1.1, 2.1, 3.1)

crps(ens, obs_ens)
```

Lower values indicate a sharper and better-centered ensemble forecast.

### Interval Evaluation with `picp()` and `mwpi()`

`picp()` and `mwpi()` work with deterministic lower and upper predictive
intervals.

- `picp(lower, upper, obs)` evaluates inclusive coverage
- `mwpi(lower, upper)` evaluates average interval width

```{r}
lower <- c(0.9, 1.9, 2.9, 3.8)
upper <- c(1.3, 2.3, 3.3, 4.4)
obs_int <- c(1.1, 2.4, 3.1, 4.0)

picp(lower, upper, obs_int)
mwpi(lower, upper)
```

Coverage and width should be interpreted together:

- higher `picp()` is better when the target is reliable coverage
- lower `mwpi()` is better when the target is sharp intervals

Wide intervals can inflate coverage, so `picp()` alone is not enough.

### Relative Improvement with `skill_score()`

`skill_score()` compares a lower-is-better forecast score to a lower-is-better
baseline score using:

`1 - mean(score) / mean(baseline_score)`

```{r}
score <- c(0.8, 0.7, 0.9)
baseline <- c(1.0, 1.0, 1.0)

skill_score(score = score, baseline_score = baseline)
```

Positive values indicate improvement over the baseline. Negative values mean
the candidate score is worse than the baseline on average.

## Practical Examples

### Interval Evaluation Workflow

A simple interval workflow pairs coverage and width rather than optimizing one
in isolation.

```{r}
interval_summary <- c(
  picp = picp(lower, upper, obs_int),
  mwpi = mwpi(lower, upper)
)

interval_summary
```

Use this kind of summary when you want to compare competing interval methods
and inspect the reliability-versus-sharpness tradeoff explicitly.

### Ensemble Evaluation Workflow

For ensemble forecasts, `crps()` is the most direct current package entry
point.

```{r}
better_ens <- ens
worse_ens <- ens + 0.8

c(
  better = crps(better_ens, obs_ens),
  worse = crps(worse_ens, obs_ens)
)
```

This is often a better fit than forcing the ensemble into a single
deterministic summary before evaluation.

### Skill Score Workflow

`skill_score()` is useful when you already have forecast scores from two
comparable methods and want a simple relative-improvement summary.

```{r}
candidate_score <- c(0.6, 0.7, 0.8)
reference_score <- c(0.9, 0.9, 0.9)

skill_score(candidate_score, reference_score)
```

The result depends directly on the baseline choice, so the baseline must be
defensible and stable.

## Deterministic vs Special-Interface Boundaries

`gof()` is still the right entry point for deterministic single-series
evaluation:

```{r}
gof(sim, obs)
```

But the probabilistic wrappers are intentionally separate because their input
contracts differ:

- `crps()` needs an ensemble matrix
- `picp()` needs lower and upper interval vectors plus observations
- `mwpi()` needs lower and upper interval vectors
- `skill_score()` needs score values and baseline score values

That is why `gof(sim, obs, extended = TRUE)` does not auto-select them:

```{r}
intersect(
  c("crps", "picp", "mwpi", "skill_score"),
  names(gof(sim, obs, extended = TRUE))
)
```

## Caveats and Limits

- `hydroMetrics` does not currently provide a full probabilistic calibration
  framework.
- The uncertainty-facing metrics are wrapper-driven and contract-specific.
- Interval diagnostics need both coverage and width to be interpreted well.
- `skill_score()` is only as meaningful as the baseline it compares against.
- Deterministic metrics and probabilistic metrics should not be treated as
  interchangeable summaries of the same objective.

Choose the evaluation path that matches the forecast product you actually have:

- single deterministic series: `gof()` or deterministic wrappers
- ensemble forecasts: `crps()`
- predictive intervals: `picp()` and `mwpi()`
- score comparison against a reference: `skill_score()`

## Related Guides

For deterministic calibration workflows, see:

- `vignette("calibration-guide", package = "hydroMetrics")`
- `vignette("getting-started", package = "hydroMetrics")`

For exact function contracts, consult:

- `?gof`
- `?crps`
- `?picp`
- `?mwpi`
- `?skill_score`
